---
output:
  html_document: default
  pdf_document: default
---
**Predicting Cancer Type Based On Gene Expression Levels**
_Nelson Gonzabato_
_13th October 2018_
In this Kernel, I'll explore data used to classify patient's cancer type. The data is based on  where they classified patients s either having Acute Myeloid Leukemia(AML) or Acute Lymphoblastic Leukemia.
  I decided to use the first 38 patients to train the models to predict patient cancer type based on expression levels of genes given by their accession numbers.
  
```{r}
shh<-suppressMessages
shh(library(tidyverse))
shh(library(caret))
shh(library(reshape2))
trainset<-read.csv("data_set_ALL_AML_train.csv",stringsAsFactors = F,quote = "")
head(trainset,10)
#Replace everything that starts with X as some other name. Number
#Num is the experiment number?
colnames(trainset)<-gsub("X{1,}","Num",colnames(trainset))
trainset<-trainset %>% 
 select(-contains("call"),-Gene.Description) %>% 
  melt(id.vars="Gene.Accession.Number") %>% 
  spread(Gene.Accession.Number,value) %>% 
  rename(ExptNo.=variable)
#Find Missing values
#trainset %>% 
  #map_dbl(~sum(is.na(.x)))
#That was a bad way to check for NAs for a very large dataset.
#Simply use anyNa
anyNA(trainset)
#dataset contains no missing values
#Load patient data and cancer type
actual<-read.csv("actual.csv",stringsAsFactors = F,quote="")
#Susbset data for first 72 patients
trainme<-trainset[1:72,]
#bind to training set
patientdata<-cbind(trainme,actual)
patientdata<-patientdata %>% 
  select(patient,cancer,everything()) %>% 
  mutate(patient=as.factor(patient),cancer=as.factor(cancer))
training<-patientdata %>% 
  filter(!is.na(ExptNo.)) 
#Use this data to train the models on Cancer type. 
patientdata

```
 In this part, I'll visualize the expression levels. Since the dataset is huge, we've only visualised the top 20 genes expressed. 
```{r}

patientdata %>% 
select(-ExptNo.) %>% 
  gather("id","value",3:ncol(.)) %>% 
 mutate(GeneAccNum=as.factor(id),genelvls=value) %>% 
  select(-id,-value) %>% 
  filter(!is.na(genelvls)) %>% 
  top_n(20) %>% 
 ggplot(aes(GeneAccNum))+geom_bar(aes(fill=genelvls))+
  theme(axis.text.x = element_text(angle=90),
plot.title = element_text(hjust=0.5))+facet_wrap(~cancer)+
  ggtitle("Top 20 Genes Expressed in AML and ALL patients")+
  labs(x="Gene Accession Number",caption="Based on original dataset")+
  coord_flip()+
  scale_fill_continuous(low="firebrick3",high = "hotpink")

colours()
  
```
The above plot suggests that there is some overlap in gene expression although the levels of these genes are different in AML and ALL patients. The most commonly expressed gene in AML for instance is M25079_s_at

In this part we'll apply basic machine learning to try and predict cancer type based on patient gene expression levels.
```{r}
#First partition the data into a training and validating dataset.
training<-training %>% 
         select(-patient,-ExptNo.)
traincanc<-createDataPartition(training$cancer,p=0.85,list=F)
validatecanc<-training[-traincanc,]
traincanc<-training[traincanc,]
#Setup metric and control. We'll use accuracy as our metric
control<-trainControl(method="repeatedcv",number=10,repeats = 3)
metric<-"Accuracy"
#Set up models
set.seed(233)
fit.svm1<-train(cancer~.,data=traincanc,method="svmRadial",trControl=control,metric=metric,
                verbose=F)
set.seed(233)
fit.knn1<-train(cancer~.,data=traincanc,method="knn",trControl=control,metric=metric)
set.seed(233)
fit.rf1<-train(cancer~.,data=traincanc,method="rf",trControl=control,metric=metric,
               verbose=F)
result<-resamples(list(rf=fit.rf1,svm=fit.svm1,knn=fit.knn1))
dotplot(result)
```
In the above section, we see that the knn and random forest model give us the highest accuracy. However, we'll validate it using our validator dataset to see how it works on unseen data.
```{r}
predictcanc<-predict(fit.rf1,validatecanc)
confusionMatrix(predictcanc,validatecanc$cancer)
```
The confusion matrix does suggest that our model does a pretty good job of predicting cancer types. We can therefore test this on more unseen data. The independent csv file.
```{r}
testcanc<-read.csv("data_set_ALL_AML_independent.csv")
#Process our data to look ust like the training set
head(testcanc)
#Change col headings
colnames(testcanc)<-gsub("X{1,}","Num",colnames(testcanc))
#confirm changes
head(testcanc)
#Process data
testcanc<-testcanc %>% 
  select(-contains("call"),-Gene.Description) %>% 
  melt(id.vars="Gene.Accession.Number") %>% 
  spread(Gene.Accession.Number,value) %>%  
  rename(ExptNo.=variable) %>% 
  select(-ExptNo.)

#Check for missing values
anyNA(testcanc)
#dateset contains no missing values
#Apply model
predicttest<-predict(fit.knn1,testcanc)
testcanc1<-testcanc %>% 
          mutate(cancer=as.factor(predicttest),patientId=row_number()) %>% 
          select(patientId,cancer,everything())
#Confusion matrix
confusionMatrix(predicttest,testcanc1$cancer)

```
The prediction can be compared to the actual csv file starting from patient 39. The prediction is not 100% accurate and will therefore need refinement. There could also be a problem of overfitting. 
```{r}
head(testcanc1)
```
```{r}
actual[39:nrow(actual),]
```

Thank you for viewing! 
